Large and Infinitary Quotient Inductive-Inductive Types

### 2020-01-10 draft/paper.tex
paper

\begin{definition}[The $\Set$ model of $\ToS$] For each $j$, we have $\bA : \ToS_{j+1, j}$, which can be given by restricting the $\bM$ model
of Section \ref{sec:semantics} so that we only have the first $\Con$ components
in the interpretations for contexts, substitution, types, term, and we only have
actions on contexts in the interpretations of term and substitution
formers. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bA} && :\equiv \Set_{j+2} \\
  & \Sub_{\bA}\,\Gamma\,\Delta && :\equiv \Gamma \ra \Delta\\
  & \Ty_{\bA}\,\Gamma && :\equiv \Gamma \ra \Set_{j+2}\\
  & \Tm_{\bA}\,\Gamma\,A && :\equiv (\gamma : \Gamma) \ra A\,\gamma
\end{alignat*}
Now, for $\Gamma : \Con_{\syn}$, the type of $\Gamma$-algebras is given by
$\llb\Gamma\rrb_{\bA}$. E.g.\, $\llb\NatSig\rrb_{\bA}$ yields a left-nested
$\Sigma$-type of pointed sets with an endofunction. Also, $\llb\Gamma\rrb_{\bM}$
extends $\llb\Gamma\rrb_{\bA}$ to an flCwF of $\Gamma$-algebras.
\end{definition}

\begin{definition}[Logical predicate model of $\ToS$ over the $\Set$ model]
For each $j$ level we have $\bD$, which is a displayed $\ToS$ model over $\bA$
at level $j$. This model, analogously to $\bA$, is given by restricting $\bM$ to
the \emph{third} components everywhere, corresponding to types or actions on
types. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bD}\,\ulGamma && :\equiv \ulGamma \ra \Set_{j+2} \\
  & \Sub_{\bD}\,\Gamma\,\Delta\,\ulsigma && :\equiv (\ulgamma : \ulGamma) \ra \Gamma\,\gamma \ra \Delta(\llb\ulsigma\rrb_{\bA}\,\ulgamma)\\
  & \Ty_{\bD}\,\Gamma\,\ulA && :\equiv (\ulgamma : \ulGamma) \ra \Gamma\,\ulgamma \ra \llb\ulA\rrb_{\bA}\,\ulgamma \ra \Set_{j+2}\\
  & \Tm_{\bD}\,\Gamma\,A\,\ult && :\equiv (\ulgamma : \ulGamma)(\gamma : \Gamma\,\ulGamma)
    \ra A\,\ulgamma\,\gamma\,(\llb\ult\rrb_{\bA}\,\ulgamma)
\end{alignat*}
\end{definition}

For $\Gamma : \SynSig_j$, the type of displayed $\Gamma$-algebras over some
$\ulgamma : \llb\Gamma\rrb_{\bA}$ is given by $\llb\Gamma\rrb_{\bD}\,\ulgamma$. In
other words, $\llb\Gamma\rrb_{\bD}$ yields the notion of types in the flCwF of
$\Gamma$-algebras given by $\llb\Gamma\rrb_{\bM}$.

\begin{definition}[Displayed algebra section model of $\ToS$] Analogously to
$\bA$ and $\bD$, we define $\bS$ as a displayed $\ToS$ model over the total
model of $\bD$, which is given by restricting $\bM$ to the fourth components,
corresponding to interpretations of terms and actions on terms.

For $\ulgamma : \llb\Gamma\rrb_{\bA}$ and $\gamma : \llb\Gamma\rrb_{\bD}\,\ulgamma$,
the type of $\Gamma$-sections is computed as $\llb\Gamma\rrb_{\bS}\,\ulgamma\,\gamma$.
\end{definition}

@@

\emph{Remark.} If we start with a syntactic signature at level $j$, then the
underlying sets in the term algebra are all in $\Set_{j+1}$. Hence, the term
algebra for $\NatSig : \SynSig_0$ has an underlying set in $\Set_{j+1}$. This is
a bit inconvenient, since normally we would have natural number in $\Set_0$. Our
current term model construction cannot avoid this level bump, since $\syn_j$ is
necessarily a large type, and we do not have a way to construct a small set from
a large set of terms. Perhaps this would be possible with a \emph{resizing rule}
\cite{voevodsky2011resizing}. Also, if we only consider closed finitary QIITs,
with no possibility of referring to external types in signatures, then we can
modify the current term model construction so that we always build sets in
$\Set_0$. This would cover natural numbers and most dependent type theories.

@@

%% We assume $\ulOmega : \SynSig_j$.
%% We abbreviate the term algebra
%% $\llb\ulOmega\rrb_{\bT_{\ulOmega}}\,\id$ as

%% $\llb\ulOmega\rrb_{\bT_{\ulOmega}}\,\id$ as b
%% $\llb\ulOmega\rrb_{\bT_{\ulOmega}}\,\id$ as

@@

Cartmell \cite{gat} defines generalised algebraic theories (GATs)
using type theoretic syntax. Compared to our QII signatures, his
signatures support sort equations but do not cover infinitary
constructors or recursive equations. A way to encode sort equations in
our signatures is using isomorphisms instead of equalities. It seems
that infinitary constructors and sort equations are not compatible in
a setting with UIP. In contrast to our algebraic definition,
Cartmell's signatures are given by presyntax, named variables and
typing relations. He also does not construct categories of algebras
explicitly, but is concerned with showing that the classifying
category for GATs is that of contextual categories.

A more semantic approach to QIITs is given by Altenkirch et al.\
\cite{gabe}. They generalise the usual functorial semantics of
inductive types to QIITs by considering towers of functors and
building complete categories of algebras from such towers. Their
notion of signature does not enforce strict positivity, hence
describes a larger class of QII signatures. The equivalence of
initiality and induction is showed, but initial algebras are not
constructed.

Kaposi et al's work \cite{kaposi2019constructing} is the most direct
precursor of our work. They don't consider infinitary constructors or
constructors with recursive equations which make their semantics much
simpler.

@@

Higher inductive types (HITs) are generalisations of QIITs in settings
with proof-relevant identity types, in fact, they were introduced
before QIITs \cite{HoTTbook}. \cite{niels} is a description of a
subset of HITs. \cite{hiit} describes a syntax for higher
inductive-inductive types using a theory of signatures similar to ours,
however does not construct categories of algebras or show
initiality. Semantics for different subclasses of HITs are given by
\cite{lumsdaineShulman,10.1145/3209108.3209130,nielsmsc,sojakova,moeneclaey}. Cubical
type theories were shown to support HITs in a computational way
\cite{niels,hiit}.

Our notion of displayed CwF is an extension of displayed category
\cite{displayedcats} however in a setting with UIP.

\section{Conclusions and further work}
>>>>>>> eba72f32c6607b337c6315a124afe03f6f3ad92d
\label{sec:conclusion}

>>>>

\begin{definition}[The $\Set$ model of $\ToS$] For each $i$ and $j$, we have $\bA : \ToS_{\max(i+1, j)+1, j}$, which can be given by restricting the $\bM_{i,j}$ model
of Section \ref{sec:semantics} so that we only have the first $\Con$ components
in the interpretations for contexts, substitution, types, term, and we only have
actions on contexts in the interpretations of term and substitution
formers. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bA} && :\equiv \Set_{\max(i+1, j)} \\
  & \Sub_{\bA}\,\Gamma\,\Delta && :\equiv \Gamma \ra \Delta\\
  & \Ty_{\bA}\,\Gamma && :\equiv \Gamma \ra \Set_{\max(i+1, j)}\\
  & \Tm_{\bA}\,\Gamma\,A && :\equiv (\gamma : \Gamma) \ra A\,\gamma
\end{alignat*}
Now, for some $\Gamma : \Con_{\syn_j}$, the type of $\Gamma$-algebras at level
$i$ is given by $\llb\Gamma\rrb_{\bA}$, where we implicitly lift $\syn_j :
\ToS_{j+1, j}$ to $\ToS_{\max(i+1,j)+1, j}$. E.g.\, $\llb\NatSig\rrb_{\bA}$
yields a left-nested $\Sigma$-type of pointed sets with an endofunction. Also,
$\llb\Gamma\rrb_{\bM}$ extends $\llb\Gamma\rrb_{\bA}$ to an flCwF of
$\Gamma$-algebras.
\end{definition}

\begin{definition}[Logical predicate model of $\ToS$ over the $\Set$ model]
For each $i$ and $j$ level we have $\bD$, which is a displayed $\ToS$ model over
$\bA$. This model, analogously to $\bA$, is given by restricting $\bM_{i,j}$ to
the \emph{third} components everywhere, corresponding to types or actions on
types. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bD}\,\ulGamma && :\equiv \ulGamma \ra \Set_{\max(i+1, j)} \\
  & \Sub_{\bD}\,\Gamma\,\Delta\,\ulsigma && :\equiv (\ulgamma : \ulGamma) \ra \Gamma\,\gamma \ra \Delta(\llb\ulsigma\rrb_{\bA}\,\ulgamma)\\
  & \Ty_{\bD}\,\Gamma\,\ulA && :\equiv (\ulgamma : \ulGamma) \ra \Gamma\,\ulgamma \ra \llb\ulA\rrb_{\bA}\,\ulgamma \ra \Set_{\max(i+1, j)}\\
  & \Tm_{\bD}\,\Gamma\,A\,\ult && :\equiv (\ulgamma : \ulGamma)(\gamma : \Gamma\,\ulGamma)
    \ra A\,\ulgamma\,\gamma\,(\llb\ult\rrb_{\bA}\,\ulgamma)
\end{alignat*}
\end{definition}

For $\Gamma : \SynSig_j$, the type of displayed $\Gamma$-algebras at level $i$
over some $\ulgamma : \llb\Gamma\rrb_{\bA}$ is given by
$\llb\Gamma\rrb_{\bD}\,\ulgamma$. Here, we also implicitly lift $\Gamma$ to live
in the appropriately sized $\syn$. In other words, $\llb\Gamma\rrb_{\bD}$ yields
the notion of types in the flCwF of $\Gamma$-algebras given by
$\llb\Gamma\rrb_{\bM}$.

\begin{definition}[Displayed algebra section model of $\ToS$] Analogously to
$\bA$ and $\bD$, for each $i$ and $j$ levels we define $\bS$ as a displayed
  $\ToS$ model over the total model of $\bD$, which is given by restricting
  $\bM_{i,j}$ to the fourth components, corresponding to interpretations of terms and
  actions on terms.

For $\ulgamma : \llb\Gamma\rrb_{\bA}$ and $\gamma :
\llb\Gamma\rrb_{\bD}\,\ulgamma$, the type of $\Gamma$-sections at level $i$ is
computed as $\llb\Gamma\rrb_{\bS}\,\ulgamma\,\gamma$.
\end{definition}

@@

\emph{Remark.} If we start with a syntactic signature at level $j$, then the
underlying sets in the term algebra are all in $\Set_{j+1}$. Hence, the term
algebra for $\NatSig : \SynSig_0$ has an underlying set in $\Set_{j+1}$. This is
a bit inconvenient, since normally we would have natural number in $\Set_0$. Our
current term model construction cannot avoid this level bump, since $\syn_j$ is
necessarily large, and we do not have a way to construct a small set from
a large set of terms. Perhaps this would be possible with a \emph{resizing rule}
\cite{voevodsky2011resizing}. Also, if we only consider closed finitary QIITs,
with no possibility of referring to external types in signatures, then we can
modify the current term model construction so that we always build sets in
$\Set_0$. This would cover natural numbers and most dependent type theories.

\subsection{Cumulativity of Algebras}

We would like to show that term algebras are initial, but we want to do this
\emph{on all universe levels}, i.e.\ that terms algebras are initial when lifted
to any higher level. This requires showing that $\Gamma$-algebras are
cumulative. We do this by induction on syntactic signatures.

\begin{definition}[Cumulativity model] We assume $j$, $k$ and $l$ levels such that $j+1 \leq k$, $j+1 \leq l$ and $k \leq l$. We define a displayed model over $\syn_j : \ToS_{j+1, j}$ lifted to
$\ToS_{l, j}$. In the following, we notate the level of algebras computed by
  $\llb\blank\rrb_{\bA}$ as an extra index, as in
  $\llb\ulGamma\rrb_{\bA_k}$. The underlying sets of the model are as follows.
\begin{alignat*}{3}
  & \Con\,\ulGamma && :\equiv \Subtype\,\llb\ulGamma\rrb_{\bA_k}\,\llb\ulGamma\rrb_{\bA_l}\\
  & \Sub\,\Gamma\,\Delta\,\ulsigma && :\equiv (\gamma : \llb\ulGamma\rrb_{\bA_k})\ra
  \llb\ulsigma\rrb_{\bA_k}\,\gamma = \llb\ulsigma\rrb_{\bA_l}\,\gamma\\
  & \Ty\,\Gamma\,\ulA &&:\equiv (\gamma : \llb\ulGamma\rrb_{\bA_k})\ra
      \Subtype\,(\llb\ulA\rrb_{\bA_k}\,\gamma)\,(\llb\ulA\rrb_{\bA_l}\,\gamma)\\
  & \Tm\,\Gamma\,A\,\ult &&:\equiv (\gamma : \llb\ulGamma\rrb_{\bA_k})\ra
     \llb\ult\rrb_{\bA_k}\,\gamma = \llb\ult\rrb_{\bA_l}\,\gamma
\end{alignat*}
The rest of the model is straightforward to define. Now, it follows from the
induction assumption for $\syn$ and the reflection rule for $\Subtype$ in
Section \ref{sec:cumulativity}, that $\llb\ulGamma\rrb_{\bA_k} \leq
\llb\ulGamma\rrb_{\bA_l}$.

@@

@@

Cartmell \cite{gat} defines generalized algebraic theories (GATs) using type
theoretic syntax. Compared to our QII signatures, his signatures support sort
equations but do not cover infinitary constructors or recursive equations. A way
to encode sort equations in our signatures is using isomorphisms instead of
equalities. In contrast to our algebraic definition, Cartmell's signatures are
given by presyntax, named variables and typing relations, there is no
explicit model theory provided for signatures, and no explicit term model
construction. Cartmell focuses instead on showing that contextual categories
serve as classifying categories for GATs.

A more semantic approach to QIITs is given by Altenkirch et al.\
\cite{gabe}. They generalize the initial algebra semantics of
inductive types to QIITs by considering towers of functors and
building complete categories of algebras from such towers. Their
notion of signature does not enforce strict positivity, hence
describes a larger class of QII signatures. The equivalence of
initiality and induction is showed, but initial algebras are not
constructed.

Kaposi et al's work \cite{kaposi2019constructing} is the direct precursor of our
work. They don not consider infinitary constructors or constructors with
recursive equations, which make their semantics considerably simpler. They also
do not provide a model theory of signatures, instead they assume signatures
as an ad-hoc QIIT.

@@

Higher inductive types (HITs) are generalisations of QIITs in settings with
proof-relevant identity types. They were introduced before QIITs
\cite{HoTTbook}. \cite{niels} is a description of a subset of HITs. \cite{hiit}
describes a syntax for higher inductive-inductive types using a theory of
signatures similar to ours, however does not construct categories of algebras or
show initiality. Semantics for different subclasses of HITs are given by
\cite{lumsdaineShulman,10.1145/3209108.3209130,nielsmsc,sojakova,moeneclaey}. Cubical
type theories were shown to support HITs in a computational way
\cite{niels,hiit}.

Our notion of displayed CwF is an extension of displayed categories
\cite{displayedcats}, although in a setting with UIP.

\section{Conclusions and further work}
\label{sec:conclusion}
